---
title: "Big Data and Machine Learning for transport data analysis: A tutorial"
author: "Robin Lovelace and Liam Bolton"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{mlCars-tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction
This tutorial will offer a brief overview of what you can do with the [mlCars](https://github.com/Robinlovelace/mlCars/) package. Here, we focus on exploring the data, performing regression analysis and some machine learning. To find out more about the DfT-funded mlCars project, read this [report](https://github.com/Robinlovelace/mlCars/tree/master/docs)

To reproduce the results demonstrated in this tutorial we assume that:
- You have some experience with programming in general or are not completely new to R. If you are new to R, we recommend you carry out a a few hours of this free [introductory course](https://www.datacamp.com/courses/free-introduction-to-r) from [Datacamp](https://www.datacamp.com/courses/free-introduction-to-r).
- Up-to-date versions of R and RStudio are installed on your system. See online resources such as [Efficient R Programming](https://csgillespie.github.io/efficientR/set-up.html#installing-r) [@gillespie_efficient_2016] for more info.

## Data
A note about the **data**: the input data for this tutorial is an origin-destination matrix of travel patterns derived from the 2011 census and augmented with additional variables from open census and OSM data. These are represented as spatial lines (of class `SpatialLinesDataFrame` but which could be exported to a range of geographic data formats). It is not the purpose of this tutorial to explain how such datasets can be generated. This is covered in the [vignette](https://cran.r-project.org/web/packages/stplanr/vignettes/introducing-stplanr.html) of the [stplanr R package](https://github.com/ropensci/stplanr).

From your RStudio session, the necessary packages and data can be installed with the following commands. This will install the requisite **devtools** and **mlCars** packages:

```{r, eval=FALSE}
install.packages("devtools")
devtools::install_github("robinlovelace/mlCars")
```

The **mlCars** package contains the example data used in this tutorial. Load it with the `library()` command:

```{r}
library(mlCars)
library(stplanr)
library(dplyr)
```

The main input files resulting from the preprocessing stages is `wyflows`, which can be loaded with the following command:

```{r, eval=FALSE}
wyflows = readRDS("data/wyflows.Rds")
```

The first thing to do with this data is to gain a basic understanding of it:

```{r}
class(wyflows)
```

Let's have a look at the variables

```{r}
names(wyflows)
```

Inspect the last few rows

```{r}
tail(wyflows)
```

Take a look at some summary stats

```{r}
summary(wyflows$car)
```

Let's create some quick visualisations in base R. First, a boxplot showing the distribution of car distance counts

```{r}
boxplot(wyflows$car)
```

And... a bar chart
```{r}
counts = table(wyflows$car)
barplot(counts, xlab='', ylab='', ylim = c(0,14000), main="Car")
```

How about a static map to give our visualisations a spatial element?
```{r, echo=FALSE}
wyflows = spTransform(wyflows, CRS("+init=epsg:4326"))
wyflows = wyflows[wyflows$distance > 0,]
library(tmap)
qtm(wyflows[wyflows$npeople > 200,], lines.col = "car", lines.style = "pretty")
```

Let's do some regression analysis. First, we need to do some data cleaning

Run a linear model predicting car use as a function of explanatory variables:

```{r}
m_age = lm(
  car ~ `25-34` + `35-49` + `50-64` + `65-74` + `75+`,
  data = wyflows@data)
summary(m_age)
```

We can look at the links with other variables using the same approach.
The code below, for example, shows a very slight positive correlation with central heating:

```{r}
mCentheat = lm(car ~ centheat, data = wyflows@data)
summary(mCentheat)
```


### Machine Learning 

Now that we've covered visualisation and traditional statistical analysis, let's do some machine learning! We've used [XGBoost](http://xgboost.readthedocs.io/), an accessible package for fitting boosted trees, to demonstrate how to build a machine learning model using transport data from the **mlCars** package.

```{r}
# install.packages("xgboost") # install if needed
library(xgboost)
```

Set seed for reproducibility
```{r}
set.seed(2017)
```

Select the variables to use in the training dataset and convert to numeric, on subset of data with > 20 car drivers:

```{r}
flows = wyflows[wyflows$car > 100,]
allmodes = select(flows@data, metro:othertransp) %>% 
  mutate_all(function(x) x / rowSums(.)) %>% 
  select(-car)
```

Build a model, focusing on metro transport. In the analysis we focused on the car, however
```{r}
xg = xgboost(data = as.matrix(allmodes), label = flows$car / nrow(flows), nrounds = 10,  max_depth = 7, weight = flows$car)
```

Make a prediction
```{r}
xg_fitted = predict(xg, as.matrix(allmodes))
```

Calculate how well the model has performed
```{r}
cor(xg_fitted,  flows$car / nrow(wyflows))
```

Plot the variable importance
```{r}
importance_xgb = xgb.importance(model = xg, feature_names = names(allmodes))
xgb.plot.importance(importance_xgb)
```

## Conclusion
We've covered some of the things you can do with the [**mlCars**](https://github.com/Robinlovelace/mlCars/) package including: visualisation, traditional statistical analysis and machine learning. The analysis here has not been used in the mlCars report, it is simply used to demonstrate some of the applications. If you'd like to learn more about the mlCars package you can read the report in the [vignettes]https://github.com/Robinlovelace/mlCars/blob/master/vignettes/) folder. You can also access more code from the repository at [`preprocess.R`](https://github.com/Robinlovelace/mlCars/blob/master/vignettes/preprocess.R) and [`geo.R`](https://github.com/Robinlovelace/mlCars/blob/master/vignettes/geo.R).
